---
title: "Course Material"
author: "Federica"
date: "12/4/2019"
output: pdf_document
---

# Essentials information contained in R programming coursera cours

## Objects
R has five basic or "atomic" classes of objects: 

- character
- numeric (real numbers)
- integer
- complex
- logical (True/False)

Empty vectors can be created with the function:

```{r}
vector()
```

## Attributes

R objects can have attributes:

- names, dimnames
- dimensions (e.g. matrices, arrays)
- class
- length
- other user-defined attributes/metadata

Attributes of an object ca be accessed using the attributes() function

## Creating Vectors

The c() function can be used to create vectors of objects:

```{r}
x <- c(0.5, 0.6)        ## numeric
x <- c(TRUE, FALSE)     ## logical
x <- c(T, F)            ## logical
x <- c("a", "b", "c")   ## character
x <- 9:29               ## integer
x <- c(1+0i, 2+4i)      ## complex
```

If you try to mix objects in a vector, *coercion* occurs so that every element in the vector is of the same class:

```{r}
y <- c(1.7, "a")    ##character
y <- c(TRUE, 2)     ##numeric
y <- c("a", TRUE)   ##character
```

You can also explicit coercion using the as.* functions

```{r}
x <- 0:6
class(x)
```
```{r}
as.numeric(x)
```
```{r}
as.logical(x)
```
```{r}
as.character(x)
```

## Lists

Lists are a special type of vector that can contain elements of different classes.

```{r}
x <- list(1, "a", TRUE, 1+4i)
x
```

## Matrices

Matrices are vectors with a *dimension* attribute. The dimension attribute is itself an integer vector of length 2 (nrow, ncol)

```{r}
m <- matrix(nrow = 2, ncol = 3)
m
```
```{r}
dim(m)
```
```{r}
attributes(m)
```

Matrices are constructed *column-wise*, so entries can be thought of starting in the "upper left" corner and running down the columns:

```{r}
m <- matrix(1:6, nrow = 2, ncol = 3)
m
```

Matrices can also be created directly from vectors by adding a dimension attribute:

```{r}
m <- 1:10
m
```
```{r}
dim(m) <- c(2,5)
m
```

Furthermore, matrices can be created by *column-binding* or *row-binding* with cbind() and rbind()

```{r}
x <- 1:3
y <- 10:12
cbind(x, y)
```
```{r}
rbind(x, y)
```

